\ripple: Differentiable Waveforms for Gravitational Wave Data Analysis

### 2022-12-28 src/tex/ms.tex

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}

% Author list
\author{Adam Coogan} \UdeM \Mila
\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
\author{Daniel Foreman-Mackey} \CCA
\author{Maximiliano Isi} \CCA
\author{Kelvin Lam} \CCA
\author{Kaze W.~K.~Wong} \CCA
\author{Aaron Zimmerman} \CGP

@@

% \te{Kelvin should add stuff here}

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{../static/amplitude_loss.pdf}
    \caption{Relative error of amplitudes of IMRPhenomD waveforms. Blue: Error
        between original Phenom waveform and NR waveform; Orange: Error between
        optimized waveform and NR waveform. \te{Can you add the data and the plotting script to the repo so that we can generate the plots using the same matplotlib settings?}}
    \label{fig:loss_compare}
    \end{figure}

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are commonly
regarded as the most accurate, they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating "approximants" of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
are not as accurate but can be evaluated much faster. There are a few families
of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate
model. \kw{cite and add a couple more sentences to introduce the waveform
families.} \te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}

@@

% Motivation
Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Individual segments are then connected together such that they are
differentiable. Using this method, the generated waveforms are expected to be
erroneous, as the connections would affect the originally fitted segments. The
coefficients of the current IMRPhenomD implementation are often hand tuned, and
are tuned in sub-sets of parameters instead of all together. This means the
tuning process ignore the correlation between different sub-sets of parameters,
so the best-fitted solution obtained by fitting sub-sets of parameters may not
be the global optimum compared to fitting all coefficients at once. Instead, by
jointly fitting all coefficients, the inaccuracy in fitting individual and
connecting different segments can be accounted during the optimization.
Therefore, there is potential accuracy one can gain by fitting all coefficients
at once.

\kw{Kelvin Fill}
Since we can differentiate through the entire waveform model against the fitting
parameters, one can use gradient descent to find the optimal fitting parameters
all at once instead of fitting them in subsets.

Here, we define the mismatch as
\begin{equation}
	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}), 
\end{equation} 
In our optimization procedure, we took NR waveforms stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients. 
By defining the mean of individual mismatches as the loss function for optimization, we can calibrate the set of coefficients using gradient descent.

\kw{Describe figure losscompare}

\kw{Describe figure logdiff}
As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
In general, the error of optimized waveform is lower than that of the original waveform. 
The error in the early inspiral region is substantially decreased by half and other regions also show good improvement in accuracy.

\begin{figure}[t!]
\centering
\includegraphics[width=\linewidth]{../static/MM_logdiff_q148_LIGO.pdf}
\caption{Distribution of the log difference of mismatches between optimized
    and original model. The mean shows a 36\% decrease in mismatch. }
\label{fig:MM_logdiff}
\end{figure}

Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
In Fig. \ref{fig:logdiff_q148}, the distribution of log difference of mismatches are shown. 
The distribution mean lies towards the negative side, which indicates a improvement in the model. 
Despite there are testing waveforms that performed worse after optimization, it is verified that the ansatz of IMRPhenomD does not fit well with most of these waveforms, hence cannot be improved by optimization.  
Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.

\kw{Describe implication}
% Implication

This fine-tuning step is rather simple and general. Given a different
differentiable waveform model, one can apply the same procedure to optimize the
coefficients in the new model, i.e. applying gradient descent to optimize all
the fitting coefficients all together. This provides a more flexible control
during the calibration process, and potentially better final result.
With further modifications to the optimization process, the set of waveform
coefficients can be retuned to give almost an order-of-magnitude improvement.
\kw{Where is the proof for this?}

>>>>

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}

% Author list
\author{Adam Coogan} \UdeM \Mila
\author{Thomas D.~P.~Edwards} \JHU \OKC \NORDITA
\author{Daniel Foreman-Mackey} \CCA
\author{Maximiliano Isi} \CCA
\author{Kelvin K.~H.~Lam} \CCA \CUHK
\author{Kaze W.~K.~Wong} \CCA
\author{Aaron Zimmerman} \CGP

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are commonly
regarded as the most accurate, they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating "approximants" of the
full numerical relativity waveforms, which are essentially simpler ansatzes that

are not as accurate but can be evaluated much faster. 
There are a few families of approximants, such as the IMRPhenom family, SEOB family, and the NRSurrogate model. 
\kw{cite and add a couple more sentences to introduce the waveform families.} 
\te{I think a full dicussion of the various waveform families just belongs in section 2, so don't need too much more here I'd say.}

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Individual segments are then connected together such that they are
differentiable. Using this method, the generated waveforms are expected to be
erroneous, as the connections would affect the originally fitted segments. The
coefficients of the current IMRPhenomD implementation are often hand tuned, and
are tuned in sub-sets of parameters instead of all together. This means the
tuning process ignore the correlation between different sub-sets of parameters,
so the best-fitted solution obtained by fitting sub-sets of parameters may not
be the global optimum compared to fitting all coefficients at once. Instead, by
jointly fitting all coefficients, the inaccuracy in fitting individual and
connecting different segments can be accounted during the optimization.
Therefore, there is potential accuracy one can gain by fitting all coefficients
at once.

Since we can differentiate through the entire waveform model against the fitting
parameters, one can use gradient descent to find the optimal fitting parameters
all at once instead of fitting them in subsets.

Here, we define the mismatch and loss function as
\begin{equation}
	\mathcal{M}=1-\mathrm{m}(h_{\mathrm{IMR}}|h_{\mathrm{NR}}).
\end{equation}
In the optimization procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa} as training waveforms to fit waveform coefficients.  
Then, combining individual mismatches, a loss function is constructed as follows:
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i, 
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$ is the total number of training waveforms used in optimization. 
Then, one can proceed to perform optimization on waveform coefficients.

\begin{figure}[t]
	\script{loss_compare.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/loss_compare.pdf}
		\caption{
			Relative error of IMRPhenomD waveform amplitudes. Blue: Error
			between original Phenom waveform and NR waveform; Orange: Error between
			optimized waveform and NR waveform.
		}
		\label{fig:loss_compare}
	\end{centering}
\end{figure}

As demonstration, Fig. \ref{fig:loss_compare} shows the relative error of original and optimized waveform at different frequencies. 
In general, the error of optimized waveform is lower than that of the original waveform. 
The error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.

\begin{figure}[t]
	\script{mismatch_hist.py}
	\begin{centering}
		\includegraphics[width=\linewidth]{figures/mismatch_hist.pdf}
		\caption{
			Distributions of log mismatches of training waveforms. Blue: Original waveforms; Orange: Optimized waveforms.
		}
		\label{fig:mismatch_hist}
	\end{centering}
\end{figure}

Instead of examining individual waveforms, we investigate the general behavior of mismatches for a set of testing waveforms. 
In Fig. \ref{fig:mismatch_hist}, the distribution of log mismatches are shown. 
The entire optimized waveform distribution slightly shifted to the negative side. 
The peak of the original waveform distribution moved to the left end of the distribution. 
Both indicate that the optimization procedure improved the model. 
Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.

This fine-tuning step is rather simple and general. Given a different
differentiable waveform model, one can apply the same procedure to optimize the
coefficients in a new model, i.e. applying gradient descent to optimize all
the fitting coefficients all together. This provides a more flexible control
during the calibration process, and potentially better final result.
With further modifications to the optimization process, the set of waveform

coefficients can be retuned to give almost $50\%$ improvement in mismatch for well-behaved waveforms.
%\kw{Where is the proof for this?}
